---
title: "R Coding Exercise"
output: 
  html_document:
    toc: FALSE
editor_options: 
  chunk_output_type: console
---


## Loading and checking the data
I first loaded the needed packages - **dslabs** to access the _gapminder_ data and **tidyverse** to process the data and create plots. 

I then loaded the help page for _gapminder_ and asked for the structure, summary, and class. 

```{r Loading and checking the data, message=FALSE, warning=FALSE}
#Load the dslabs and other required packages
#dslabs is required to access the gapminder data 
library(dslabs)
#tidyverse is used in data processing and tidying and in plotting 
library(tidyverse)
# gtsummary is for making nice regression tables in HTML output
library(gtsummary)
# For regression diagnostic plots
library(ggfortify)

#Look at the help file for the gapminder data
help(gapminder)

#Get an overview of the data structure
str(gapminder)

#Get a summary of the data
summary(gapminder)

#Determine the type of object gapminder is
class(gapminder)
```

## Processing the data 

My first step was creating the new _africadata_ object by filtering the _gapminder_ dataset. I used only the rows for which the continent was **Africa**. I then checked the structure and summary of the new object. 

I then created two additional objects from _africadata_, one containing only infant mortality and life expectancy data, and the other containing only population and life expectancy data. I checked the structures and summaries of these as well. 

```{r Processing the data}
# Create a new object, "africadata," that contains only the data from African countries, by extracting only the rows from gapminder in which the continent is "Africa"
africadata <-
  gapminder %>%
  filter(continent == 'Africa') 

#Check the structure and summary of the data 
str(africadata)
summary(africadata)

#Create an object containing only infant_mortality and life_expectancy data within the africadata object 
v1 <- select(africadata, infant_mortality, life_expectancy)

#Create an object containing only population and life_expectancy data within the africadata object
v2 <- select(africadata, population, life_expectancy)

#Check the structure and summary of the two new objects  
str(v1)
summary(v1)
str(v2)
summary(v2)
```

## Plotting 

I created plots for each of the two new objects I created from _africadata_. Both had Life Expectancy as the outcome, while the first had infant mortality as the predictor and the second had the log population size as the predictor. 
```{r Plotting}
#Plot life expectancy as a function of infant mortality 
v1 %>%
  ggplot(aes(infant_mortality, life_expectancy)) +
  geom_point() +
  xlab('Infant Mortality') +
  ylab('Life Expectancy')

#Plot life expectancy as a function of the log population size 
v2 %>%
  ggplot(aes(population, life_expectancy)) +
  scale_x_continuous(trans = 'log2') +
  geom_point() +
  xlab('Population (log scale)') +
  ylab('Life Expectancy')
```

While these were scatter plots, they both had clusters of data points together that appeared almost linear. This was due to the data being collected over many different years. In order to tidy the data a little bit, the data was then restricted to one year. 

## More data processing 

In order to determine which year to use, I examined what years containing missing data (so that I would avoid using a year with missing data). 

The year 2000 was then used to go through the same process as above, in both processing and plotting data. 

```{r More data processing}
#Look at which years have missing data for infant mortality 
africadatana <-
  africadata %>%
  select(year, infant_mortality) %>%
  filter(is.na(infant_mortality))
#1960-1981 and 2016 have missing data 


#Creat new object with only the data from the year 2000 in the africadata object
africadata2000 <- 
  filter(africadata, year == 2000)
#View the structure and summary of the new africadata2000 object 
str(africadata2000)
summary(africadata2000)

#Recreate v1 and v2 objects from the africa data in only the year 2000 
#Create an object containing only infant_mortality and life_expectancy data within the africadata object 
v1_2000 <- select(africadata2000, infant_mortality, life_expectancy)

#Create an object containing only population and life_expectancy data within the africadata object
v2_2000 <- select(africadata2000, population, life_expectancy)

#Check the structure and summary of the two new objects  
str(v1_2000)
summary(v1_2000)
str(v2_2000)
summary(v2_2000)
```

## More plotting 
```{r More plotting}
#Recreate plots with year 2000 data 
###Plot life expectancy as a function of infant mortality 
v1_2000 %>%
  ggplot(aes(infant_mortality, life_expectancy)) +
  geom_point()

###Plot life expectancy as a function of the log population size 
v2_2000 %>%
  ggplot(aes(population, life_expectancy)) +
  scale_x_continuous(trans = 'log2') +
  geom_point()
```

## A simple fit 

The data was now much more readable. There certainly appeared to be an association between infant mortality and life expectancy, but not much of one (if any) between population size and life expectancy. To check these relationships, two simple linear models were fit, one to each relationship. 

```{r A simple fit}
#Fit linear models for the year 2000 data, with life expectancy as the outcome and infant mortality and population size as the predictors (separately), and save

## Linear model with infant mortality predictor and life expectancy outcome
fit1 <- lm(life_expectancy ~ infant_mortality, data = v1_2000)

## Linear model with (log) population size predictor and life expectancy outcome 
fit2 <- lm(life_expectancy ~ population, data = v2_2000)

#View the summaries of the two linear models 
summary(fit1)
summary(fit2)
```

Based on the outcomes of the two models, I could conclude that increased infant mortality has a negative impact on life expectancy. That is, as infant mortality increases, life expectancy decreases (p < 0.00001). However, there did not appear to be a significant relationship between life expectancy and population size (p = 0.616). 

# Predicting life expectancy (by Zane!)

What I want to do in this section is fit a basic model to predict life expectancy by year.

## EDA

The first thing I want to do here is make a spaghetti plot of how each country's life expectancy changes over the years. 

```{r}
africadata |>
  ggplot(aes(year, life_expectancy)) +
  geom_line(aes(color = country), alpha = 0.5, show.legend = FALSE) +
  geom_smooth(method = "gam", color = "black") +
  xlab('Infant Mortality') +
  ylab('Life Expectancy') +
  theme_bw()
```

Hmm, it looks like this relationship is not monotonic for several countries, and while there is a trend upwards overall, it is not necessarily linear. There is also a significant amount of variation in the trend, which is pretty normal for time series. There doesn't appear to be any seasonal/cyclic change, at least on the annual scale of measurement for which we have data.

Let's color by region and see if we see any similar trends that could explain part of the country-level effect.

```{r}
africadata |>
  ggplot(aes(year, life_expectancy)) +
  geom_line(aes(color = country), show.legend = FALSE, alpha = 0.5) +
  geom_smooth(method = "gam", color = "black") +
  xlab('Infant Mortality') +
  ylab('Life Expectancy') +
  theme_bw() +
  facet_wrap(vars(region))
```

Let's make a few more plots to see if we can observe any other relationships with life expectancy over time.

```{r}
africadata |>
  mutate(`log10 population` = log10(population), `log10 gdp` = log10(gdp),
         .keep = "unused") |>
  pivot_longer(c(infant_mortality, fertility, `log10 population`,
                 `log10 gdp`)) |>
  ggplot(aes(year, value, color = life_expectancy, group = country)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c(option = "plasma") +
  theme_bw() +
  coord_cartesian(expand = FALSE) +
  facet_wrap(vars(name), scales = "free_y")
```

I won't try and make the argument that this is the best possible visualization for these data, but I think it is good enough to give us an idea of trends. Clearly, fertility and infant mortality vary with life expectancy across time. We can see this from how the curves trend over time, and how the pattern of colors shifts over time. However, while gdp and population size appear to vary with time, I do not think that they necessarily vary with life expectancy in the same way.

Now since this is time series data, I think we are kind of obligated to plot the autocorrelation, so let's make a plot of the partial autocorrelation (as this controls for previous lags, unlike the regular autocorrelation, and is thus easier to interpret).

```{r}
africadata %>%
  select(country, year, life_expectancy) |>
  tidyr::nest(data = -country) |>
  dplyr::mutate(
    pacf_res = purrr::map(data, ~pacf(.x$life_expectancy, plot = F, lag.max = 10)),
    pacf_val = purrr::map(pacf_res, ~data.frame(lag = .x$lag, acf = .x$acf))
  ) |>
  unnest(pacf_val) |>
  ggplot(aes(x = lag, y = acf)) +
  geom_hline(yintercept = c(qnorm(0.025) / sqrt(17), qnorm(0.975) / sqrt(17)),
             lty = 2, color = "gray", size = 1.5) +
  geom_hline(yintercept = 0, size = 1.5) +
  geom_line(aes(group = country, color = country), show.legend = FALSE) +
  theme_bw(base_size = 20) +
  scale_x_continuous(labels = 1:10, breaks = 1:10, minor_breaks = NULL) +
  labs(x = "lag", y = "partial autocorrelation") +
  ggtitle("Partial autocorrelation of life expectancy by year for each African country")
```

The dashed gray lines on this plot represent approximate normal 95% confidence bands. We see that all countries have a partial autocorrelation at the 1st lag which lies outside of the 95% confidence band, but at the 2nd lag, only one country has a value outside of the band. Since we are testing a larging amount of countries, I think we can safely say that this one second lag value is spurious. The interpretation of a significant first lag partial autocorrelation value (and no other significant partial autocorrelations) is that our time series can be modeled as an autoregressive process of order 1, AKA an AR(1) process.

## Imputation

Most time series models have no way to deal with missing data. There are a lot of ways we could deal with this, and in this case we could likely find another data source to fill in the missing values with real information. But I am too lazy to do that. So first, let's look at how much data is actually missing.

```{r}
africadata |>
  select(!continent) |>
  summarize(across(everything(), ~mean(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "field", values_to = "pct_m") |>
  ggplot(aes(x = pct_m, y = forcats::fct_reorder(field, pct_m))) +
  geom_col() +
  geom_label(aes(label = scales::percent(round(pct_m, 2))), size = 10) +
  theme_bw(base_size = 20) +
  labs(x = "percent of records missing", y = NULL) +
  scale_x_continuous(labels = scales::percent_format())
```

Well, the most missing values are in GDP, which did not appear to be a strong predictor anyways, so we can throw that one out. And then since infant mortality and fertility both have missingness less than 10%, for this simple example I think it will be fine to impute with the median, although in a real analysis something more complex might be better to reduce bias.

```{r}
modeldata <- africadata |>
  dplyr::select(-continent, -gdp, -population) |>
  dplyr::mutate(
    across(c(infant_mortality, fertility),
           ~dplyr::if_else(is.na(.x), median(.x, na.rm = TRUE), .x)
           )
  )
```

## Simple model fitting

Now that the imputation is done, let's build a multivariable linear model. For this model we are going to ignore what we learned about the potentially autoregressive structure of the data :)

Now we have been looking at this by country previously, but I think that including 50+ regression parameters is maybe a bit excessive. So let's group by region instead, which will give us a much more manageable number of regression coefficients. I don't know which region of Africa we sould select as the reference group, so I will let R use the default (which in this case is Eastern Africa).

```{r}
fit3 <- lm(life_expectancy ~ . - country, data = modeldata)
fit3 |>
  tbl_regression() |>
  add_glance_source_note(
    label = list(df  ~ "Degrees of Freedom", sigma ~ "\U03C3"),
    fmt_fun = df ~ style_number,
    include = c(r.squared, AIC, sigma, df)
  )
```

We can see that all of the coefficients have low $p$-values, and the model has an $R^2$ of 0.745 (that is, the linear model explains 74.5% of the variance in life expectancy), which is pretty good. Next we should at least glance at the diagnostics.

```{r}
autoplot(fit3) + theme_bw()
```

From the diagnostics, we can see minor deviations from normality and some evidence of non-constant variance in the residuals, but I think, similar to what we saw before, the plots indicate that the residuals are correlated, so we truly do need to correct for correlated residuals. This means that the standard errors (and thus reported confidence intervals) of the linear models we reported are not necessarily reliable under this model. Let's plot the partial autocorrelations of the residuals for the model, this time using base R plotting for fun.

```{r}
pacf(residuals(fit3))
```

Yep, that is not ideal. There is definitely some residual autocorrelation of the residuals, though from this plot it is difficult to identify what the correlation structure is. So I think the next step would be adjusting for autocorrelation.

But I feel like this is long enough already, so I am just going to caution everyone about the weirdness that can show up with time series analysis and end this discussion here.
